---
title: "COVID-19 Data Analysis Report"
date: "October 1, 2021"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(error = TRUE, echo = TRUE, fig.align = 'center', cache = FALSE, cache.lazy = FALSE, message = FALSE, warning = FALSE, fig.pos = "h", error = FALSE, comment = NA)
```


```{r echo = FALSE}
if(!require(tidyverse)) install.packages("tidyverse")
if(!require(stringr)) install.packages("stringr")
if(!require(scales)) install.packages("scales")
if(!require(caret)) install.packages("caret")
if(!require(rpart)) install.packages("rpart")
if(!require(randomForest)) install.packages("randomForest")
if(!require(dplyr)) install.packages("dplyr")
if(!require(ggridges)) install.packages("ggridges")
if(!require(lubridate)) install.packages("lubridate")
if(!require(leaflet)) install.packages("leaflet")
if(!require(mapview)) install.packages("mapview"); webshot::install_phantomjs()
if(!require(kableExtra)) install.packages("kableExtra")
```

```{r echo = TRUE}
options(digits = 5)
```

## Load and Read Files


```{r echo = TRUE}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv", "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv", "time_series_covid19_deaths_global.csv","time_series_covid19_recovered_global.csv")
urls <- str_c(url_in,file_names)
urls
```


```{r echo = TRUE}
US_Cases <- read_csv(urls[1])
Global_Cases <- read_csv(urls[2])
US_Deaths <- read_csv(urls[3])
Global_Deaths <- read_csv(urls[4])

```


## Clean Data
```{r echo = TRUE}
Global_Cases <- Global_Cases%>%
    pivot_longer(cols= -c('Province/State', 'Country/Region', Lat, Long),
                 names_to ="date",
                 values_to = "cases") %>%
    select(-c(Lat,Long))

Global_Deaths <- Global_Deaths%>%
    pivot_longer(cols= -c('Province/State', 'Country/Region', Lat, Long),
                 names_to ="date",
                 values_to = "deaths") %>%
    select(-c(Lat,Long))


US_Cases <- US_Cases%>%
    pivot_longer(cols= -(UID:Combined_Key),
                  names_to = "date",
                  values_to = "cases") %>%
    select(Admin2:cases) %>%
    select(-c(Lat,Long_))

US_Deaths <- US_Deaths%>%
    pivot_longer(cols= -(UID:Population),
                  names_to = "date",
                  values_to = "deaths") %>%
    select(Admin2:deaths) %>%
    select(-c(Lat,Long_))


Global_Cases$date <- date(mdy(Global_Cases$date))
Global_Deaths$date <- date(mdy(Global_Deaths$date))
US_Cases$date <- date(mdy(US_Cases$date))
US_Deaths$date <- date(mdy(US_Deaths$date))
```

## Explore some of the values of the columns in Global Cases
```{r echo = TRUE}
# proportion of NA's to the number of rows in the data set
sum(is.na(Global_Cases$date))/nrow(Global_Cases)

#Explore some of the values of the columns Global Cases
min(Global_Cases$date)
max(Global_Cases$date)
min(Global_Cases$cases)
max(Global_Cases$cases)
```

## Explore some of the values of the columns in US Cases
```{r echo = TRUE}
# proportion of NA's to the number of rows in the data set
sum(is.na(US_Cases$date))/nrow(US_Cases)

#Explore some of the values of the columns in US Cases
min(US_Cases$date)
max(US_Cases$date)
min(US_Cases$cases)
max(US_Cases$cases)
```

## Explore some of the values of the columns in Global Deaths
```{r echo = TRUE}
# proportion of NA's to the number of rows in the data set
sum(is.na(Global_Deaths$date))/nrow(Global_Deaths)

#Explore some of the values of the columns Global Deaths
min(Global_Deaths$date)
max(Global_Deaths$date)
min(Global_Deaths$deaths)
max(Global_Deaths$deaths)
```





## Explore some of the values of the columns in US Deaths
```{r echo = TRUE}
# proportion of NA's to the number of rows in the data set
sum(is.na(US_Deaths$date))/nrow(US_Deaths)

#Explore some of the values of the columns US Deaths
min(US_Deaths$date)
max(US_Deaths$date)
min(US_Deaths$Population)
max(US_Deaths$Population)
min(US_Deaths$deaths)
max(US_Deaths$deaths)
```


## Joining data sets and Transforming

```{r echo = TRUE}

Global <- Global_Cases %>%
    full_join(Global_Deaths) %>%
    rename(Country_Region = 'Country/Region',
           Province_State = 'Province/State')
# Filter to cases > 0
Global <- Global %>% filter(cases > 0)

#Checking data after the join that it matches the data before we joined it      
summary(Global)

```




```{r echo = TRUE}
US <- US_Cases %>%
  full_join(US_Deaths)
 
#Checking data after the join that it matches the data before we joined it 
summary(US)

```



```{r echo = TRUE}

Global <- Global %>%
    unite("Combined_Key", c(Province_State,Country_Region),
          sep = ",",
          na.rm = TRUE,
          remove = FALSE)
```


```{r echo = TRUE}
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
```


```{r echo = TRUE}
uid <- read_csv(uid_lookup_url) %>%
    select(-c(Lat,Long_,Combined_Key,code3,iso2,iso3,Admin2))

Global <- Global %>%
    left_join(uid, by = c("Province_State","Country_Region")) %>%
    select(-c(UID, FIPS)) %>%
    select(Province_State,Country_Region, date, cases, deaths, Population, Combined_Key)

```
```{r echo = TRUE}
Global_totals<- Global %>%
    group_by(Province_State,Country_Region,date) %>%
    summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
    mutate(new_cases = cases - lag(cases) ,
           new_deaths = deaths - lag(deaths))
```

```{r echo = TRUE}
Global_totals
```


```{r echo = TRUE}
US_by_State <- US %>%
    group_by(Province_State,Country_Region,date) %>%
    summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
    mutate(deaths_per_mill = deaths * 1000000/ Population ) %>%
    select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
    ungroup()
US_by_State
```
```{r echo = TRUE}
US_totals <- US_by_State %>%
    group_by(Country_Region, date) %>%
    summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
    mutate(deaths_per_mill = deaths * 1000000/ Population ) %>%
    select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
    ungroup()

tail(US_totals)

```





## Analyzing and Visualizations

```{r echo = TRUE}

US_by_State <- US_by_State %>%
    mutate(new_cases = cases - lag(cases),
           new_deaths = deaths - lag(deaths))

US_totals <- US_totals %>%
    mutate(new_cases = cases - lag(cases),
           new_deaths = deaths - lag(deaths))

tail(US_totals)

```


```{r echo = TRUE}
#with new_cases added trends to COVID-19 in US
US_totals %>%
    filter(cases > 0) %>%
    ggplot(aes(x = date, y = new_cases)) +
    geom_line(aes(color = "new_cases")) +
    geom_point(aes(color = "new_cases")) + 
    geom_line(aes(y = new_deaths, color = "new_deaths")) +
    geom_point(aes(y = new_deaths, color ="new_deaths")) +
    scale_y_log10() +
    theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
    labs(title = "COVID-19 in US" , y = NULL)

```

```{r echo = TRUE}
US_cases <- ggplot(US_totals, 
 aes(date, as.numeric(new_cases))) +
 geom_col(fill = "blue", alpha = 0.6) + 
 theme_minimal(base_size = 14) +
 xlab(NULL) + ylab(NULL) + 
 scale_x_date(date_labels = "%Y/%m/%d")

```


```{r echo = TRUE}
US_cases + labs(title = "COVID-19 Daily Cases in the US" , y = NULL)

```


```{r echo = TRUE}
US_deaths <- ggplot(US_totals, 
 aes(date, as.numeric(new_deaths))) +
 geom_col(fill = "purple", alpha = 0.6) + 
 theme_minimal(base_size = 14) +
 xlab(NULL) + ylab(NULL) + 
 scale_x_date(date_labels = "%Y/%m/%d")

 US_deaths + labs(title = "COVID-19 Daily Deaths in the US" , y = NULL)

```

```{r echo = TRUE}
#with new_cases added trends to COVID-19 by State
state <- "California"
US_by_State %>%
    filter(Province_State == state) %>%
    filter(cases > 0) %>%
    ggplot(aes(x = date, y = new_cases)) +
    geom_line(aes(color = "new_cases")) +
    geom_point(aes(color = "new_cases")) +
    geom_line(aes(y = new_deaths, color = "new_deaths")) +
    geom_point(aes(y = new_deaths, color ="new_deaths")) +
    scale_y_log10() +
    theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
    labs(title = str_c("COVID-19 in ", state), y = NULL)
```

```{r echo = TRUE}
US_state_totals <- US_by_State %>%
    group_by(Province_State) %>%
    summarize(deaths = max(deaths), cases = max(cases), population = max(Population), 
              cases_per_thou = 1000 * cases /population, deaths_per_thou = 1000 * deaths/population) %>%
    filter(cases > 0, population > 0)

US_state_totals %>% 
    slice_min(deaths_per_thou, n = 10)
  
US_state_totals %>% 
    slice_max(deaths_per_thou, n = 10)


```

## Comparing Multiple Countries
```{r echo = TRUE}
# Now lets add in a few more countries

china <- Global_totals[Global_totals$Country_Region == 'China',]
spain <- Global_totals[Global_totals$Country_Region == 'Spain',]
UK <- Global_totals[Global_totals$Country_Region == 'United Kingdom',]

USplot <- ggplot(US_totals, 
 aes(date, as.numeric(new_cases))) +
 geom_col(fill = 'blue', alpha = 0.6) + 
 theme_minimal(base_size =12) +
 xlab(NULL) + ylab(NULL) + 
 scale_x_date(date_labels = "%Y/%m/%d")

China_US <- USplot + geom_col(data = china, 
 aes(date, as.numeric(new_cases)),
 fill='red',
 alpha = 0.5)

Ch_US_Sp <- China_US + geom_col(data = spain, 
 aes(date, as.numeric(new_cases)),
 fill='#E69F00',
 alpha = 0.4)

Chn_US_Sp_UK <- Ch_US_Sp + geom_col(data = UK, 
 aes(date, as.numeric(new_cases)),
 fill='#009E73',
 alpha = 0.9)

Chn_US_Sp_UK + labs(title= "China, US, UK, & Spain")
```


### Modeling
## SIR Model

```{r echo = TRUE}

state <- "California"
California <- US_by_State %>%
    filter(Province_State == state) %>%
    filter(cases > 0) 
    
```

```{r echo = TRUE}
#SIR Model

SIR <- function(time, state, parameters) {
  par <- as.list(c(state, parameters))
  with(par, {
    dS <- -beta * I * S/N
    dI <- beta * I * S/N - gamma * I
    dR <- gamma * I
    list(c(dS, dI, dR))
  })
}

```

```{r echo = TRUE}
#create a vector of cumulative cases

infected <- California %>%
    filter(cases >0) %>%
    pull(new_cases)

```

```{r echo = TRUE}
# Create an incrementing Day vector the same length as our cases vector
day <- 1:(length(infected))
N <- 14446515
#specify initial values for S, I, R
init <- c(S = N - infected[1], I = infected[1], R = 0)

RSS <- function(parameters) {
  names(parameters) <- c("beta", "gamma")
  out <- ode(y = init, times = day, func = SIR, parms = parameters)
  fit <- out[, 3]
  sum((infected - fit)^2)
}

```

```{r echo = TRUE}
# now find the values of beta and gamma that give the
# smallest RSS, which represents the best fit to the data.
# Start with values of 0.5 for each, and constrain them to
# the interval 0 to 1.0
library(deSolve)
optimization <- optim(c(0.5, 0.5), RSS, method = "L-BFGS-B", lower = c(0,0), upper = c(1, 1))

# check for convergence
optimization$message

# Optimization Parameters
opt_par <- setNames(optimization$par, c("beta", "gamma"))
opt_par

# Reproduction Number
R0 <- opt_par[1]/opt_par[2]
R0
```

# Prediction
```{r echo = TRUE}
# time in days for predictions
startdate <- min(California$date)
t <- 1:150
# get the fitted values from our SIR model
fittedcum <- data.frame(ode(y = init, times = t, func = SIR, parms = opt_par))
# add a Date column and join the observed incidence data
fittedcum <- fittedcum %>%
    mutate(date = as.Date(startdate) + t-1)%>%
             left_join(California %>% select(date,new_cases))
```



```{r echo = TRUE}

# plot the data
ggplot(fittedcum, aes(x = date)) +
  geom_line(aes(y = I), colour = "red") +
  geom_line(aes(y = S), colour = "black") +
  geom_line(aes(y = R), colour = "green") +
  geom_point(aes(y = new_cases), colour = "orange") +
  scale_y_continuous(labels = scales::comma) +
  labs(y = "Persons", title = "COVID-19 fitted vs observed cumulative incidence, California") +
  scale_colour_manual(name = "",
                      values = c(red = "red", black = "black", green = "green", orange = "orange"),
                      labels = c("Susceptible", "Recovered", "Observed incidence", "Infectious")) +
  scale_y_continuous(trans="log10")

```


## Bias
Some Bias is that I decided to analyze the state of California in more detail since I reside there. As for bias that may exist in the data would be in regards to how accurately cases are reported Globally and even state-wide. Thus, that is why I decided to focus on the state I currently reside in as it was of great interest to me. 